ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-10-09 pyked.tex
Update examples so they work

The \ck{} file that will be used in this example can be found in the
\texttt{tests} directory of the PyKED
repository.\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}}
Examining that file, we find the first section specifies the information about
the \ck{} file itself:
%
\begin{yamlbox}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: 0
chemked-version: 0.1.6
\end{yamlbox}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the dataset comes from the work of
\textcite{Mittal2006a}:
%r
\begin{yamlbox}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
\end{yamlbox}
%
Finally, this file contains just a single datapoint, which describes the experimental
ignition delay, initial mixture composition, initial temperature, initial pressure,
compression time, ignition type, and volume history that specifies
how the volume of the reactor varies with time, for simulating the compression
stroke and post-compression processes:
%
\begin{yamlbox}
datapoints:
- temperature:
    - 297.4 kelvin
  ignition-delay:
    - 1.0 ms
  pressure:
    - 958.0 torr
  composition:
    kind: mole fraction
    species:
      - species-name: H2
        InChI: 1S/H2/h1H
        amount:
          - 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        amount:
          - 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        amount:
          - 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        amount:
          - 0.63125
  ignition-type:
    target: pressure
    type: d/dt max
  compression-time:
    - 38.0 ms
  volume-history:
    time:
      units: s
      column: 0
    volume:
      units: cm3
      column: 1
    values:
      - [0.00E+000, 5.47669375000E+002]
      - [1.00E-003, 5.46608789894E+002]
      - [2.00E-003, 5.43427034574E+002]
      ...
\end{yamlbox}
%
The values for the \yabox|volume-history| are truncated here to save space. One application of the
data stored in this file is to perform a simulation using Cantera~\autocite{Cantera:2.3.0} to
calculate the ignition delay, including the facility-dependent effects represented in the volume
trace. All information required to perform this simulation is present in the \ck{} file, with the
exception of a chemical kinetic model for \ce{H2}\slash \ce{CO} combustion.

@@

\begin{pythonbox}
# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox|IdealGasReactor|, a \pybox|Wall| must be
installed between the reactor and environment and assigned a velocity. The \pybox|Wall| allows the
environment to do work on the reactor (or vice versa) and change the reactor's thermodynamic state;
we use a \pybox|Reservoir| for the environment because in Cantera, \pybox|Reservoir|s always have a
constant thermodynamic state and composition. Using a \pybox|Reservoir| accelerates the solution
compared to using two \pybox|IdealGasReactor|s, since the composition and state of the environment
are typically not necessary for the solution of autoignition problems. Although we do not show the
details here, a reference implementation of a class that computes the wall velocity given the volume
history of the reactor is available in CanSen~\autocite{cansen}, in the
\pybox|cansen.profile.VolumeProfile| class.
%
\begin{pythonbox}
# Retrieve the time and volume from the
# history in the datapoint
time = dp.volume_history.time
volume = dp.volume_history.volume
# Install the Wall between the
# reactor and environment
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))
\end{pythonbox}

@@

To calculate the ignition delay, we will follow the definition specified in the ChemKED file for
this experiment, where the experimentalists used the maximum of the time derivative of the pressure
to define the ignition delay. To calculate this derivative, we need to store the state variables and the composition on each time step, so we initialize several Python lists to act as storage:

@@

Finally, the problem is integrated using the \pybox|step| method of the \pybox|ReactorNet|. The
\pybox|step| method takes one timestep forward on each call, with step size determined by the CVODES
solver (CVODES is an adaptive time-stepping algorithm). On each step, we add the relevant variables
to their respective lists. The problem is integrated until a user-specified end time, in this case
\SI{50}{\milli\second}, although in principle, the user could end the simulation on any condition
they choose:

@@

This \ck{} file specifies multiple data points with some common
conditions, including a common mixture composition and common definition of
ignition delay. Therefore, a \yabox|common-properties| section is specified, followed by the
datapoints list:
%
\begin{yamlbox}
common-properties:
  composition: &comp
    kind: mole fraction
    species:
      - species-name: t-butanol
        InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
        amount:
          - 0.003333333
      - species-name: O2
        InChI:  1S/O2/c1-2
        amount:
          - 0.04
      - species-name: Ar
        InChI:  1S/Ar
        amount:
          - 0.956666667
  ignition-type: &ign
    target: OH*
    type: 1/2 max
datapoints:
  - temperature:
      - 1459 kelvin
    ignition-delay:
      - 347 us
    pressure:
      - 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature:
      - 1389 kelvin
    ignition-delay:
      - 756 us
    pressure:
      - 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
    ...
\end{yamlbox}
%
As before, we have truncated the datapoints list for space.

@@

Then, we load the ChemKED file and generate a list of initial conditions that will be mapped onto
the \pybox|run_simulation()| function. We first define a convenience function to collect the input
from a single datapoint and return a Python tuple with the conditions (tuples are lightweight groups
of data in Python). Then, we use the built-in \pybox|map| function to apply the
\pybox|collect_input| function to each of the elements in the \pybox|ck.datapoints| list: convert a
\pybox|Quantity| to some input units and return the magnitude, then use Python's list comprehension
syntax to generate a list of tuples (tuples are lightweight groups of data in Python):
%
\begin{pythonbox}
ck = ChemKED('Stranic2012-tbuoh.yaml')

@@

initial_conditions = map(collect_input, ck.datapoints)
initial_conditions = [(to_SI(dp.temperature, 'K'), to_SI(dp.pressure, 'Pa'), dp.get_cantera_mole_fraction()) for dp in ck.datapoints]
\end{pythonbox}

Finally, we create the processor \pybox|Pool| (with 12 processes) and send the jobs out to run:
%
\begin{pythonbox}
with Pool(12) as pool:
    ignition_delays = pool.starmap(run_simulation, initial_conditions)

>>>>

The \ck{} file that will be used in this example can be found in the
\texttt{tests} directory of the \pk{}
repository.\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}}
Examining that file, we find the first section specifies the information about
the \ck{} file itself:
%
\begin{yamlbox}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: 0
chemked-version: 0.1.6
\end{yamlbox}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the dataset comes from the work of
\textcite{Mittal2006a}:
%r
\begin{yamlbox}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
\end{yamlbox}
%
Finally, this file contains just a single datapoint, which describes the experimental
ignition delay, initial mixture composition, initial temperature, initial pressure,
compression time, ignition type, and volume history that specifies
how the volume of the reactor varies with time, for simulating the compression
stroke and post-compression processes:
%
\begin{yamlbox}
datapoints:
- temperature:
    - 297.4 kelvin
  ignition-delay:
    - 1.0 ms
  pressure:
    - 958.0 torr
  composition:
    kind: mole fraction
    species:
      - species-name: H2
        InChI: 1S/H2/h1H
        amount:
          - 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        amount:
          - 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        amount:
          - 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        amount:
          - 0.63125
  ignition-type:
    target: pressure
    type: d/dt max
  compression-time:
    - 38.0 ms
  volume-history:
    time:
      units: s
      column: 0
    volume:
      units: cm3
      column: 1
    values:
      - [0.00E+000, 5.47669375000E+002]
      - [1.00E-003, 5.46608789894E+002]
      - [2.00E-003, 5.43427034574E+002]
      ...
\end{yamlbox}
%
The values for the \yabox|volume-history| are truncated here to save space. One application of the
data stored in this file is to perform a simulation using Cantera~\autocite{Cantera:2.3.0} to
calculate the ignition delay, including the facility-dependent effects represented in the volume
trace. All information required to perform this simulation is present in the \ck{} file, with the
exception of a chemical kinetic model for \ce{H2}\slash \ce{CO} combustion.

@@

\begin{pythonbox}
# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox|IdealGasReactor|, a \pybox|Wall| must be
installed between the reactor and environment and assigned a velocity. The \pybox|Wall| allows the
environment to do work on the reactor (or vice versa) and change the reactor's thermodynamic state;
we use a \pybox|Reservoir| for the environment because in Cantera, \pybox|Reservoir|s always have a
constant thermodynamic state and composition. Using a \pybox|Reservoir| accelerates the solution
compared to using two \pybox|IdealGasReactor|s, since the composition and state of the environment
are typically not necessary for the solution of autoignition problems. Although we do not show the
details here, a reference implementation of a class that computes the wall velocity given the volume
history of the reactor is available in CanSen~\autocite{cansen}, in the
\pybox|cansen.profiles.VolumeProfile| class.
%
\begin{pythonbox}
# Retrieve the time and volume from the history in the datapoint
time = dp.volume_history.time.magnitude
volume = dp.volume_history.volume.magnitude
keywords = {'vproTime': time, 'vproVol': volume}
# Install the Wall between the reactor and environment
ct.Wall(reac, env, velocity=VolumeProfile(keywords))
\end{pythonbox}

@@

To calculate the ignition delay, we will follow the definition specified in the \ck{} file for
this experiment, where the experimentalists used the maximum of the time derivative of the pressure
to define the ignition delay. To calculate this derivative, we need to store the state variables and the composition on each time step, so we initialize several Python lists to act as storage:

@@

Finally, the problem is integrated using the \pybox|step| method of the \pybox|ReactorNet|. The
\pybox|step| method takes one timestep forward on each call, with step size determined by the CVODES
solver (CVODES uses an adaptive time-stepping algorithm). On each step, we add the relevant
variables to their respective lists. The problem is integrated until a user-specified end time, in
this case \SI{50}{\milli\second}, although in principle, the user could end the simulation on any
condition they choose:

@@

This \ck{} file specifies multiple data points with some common
conditions, including a common mixture composition and common definition of
ignition delay. Therefore, a \yabox|common-properties| section is specified, followed by the
\yabox|datapoints| list:
%
\begin{yamlbox}
common-properties:
  composition: &comp
    kind: mole fraction
    species:
      - species-name: t-butanol
        InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
        amount:
          - 0.003333333
      - species-name: O2
        InChI:  1S/O2/c1-2
        amount:
          - 0.04
      - species-name: Ar
        InChI:  1S/Ar
        amount:
          - 0.956666667
  ignition-type: &ign
    target: OH*
    type: 1/2 max
datapoints:
  - temperature:
      - 1459 kelvin
    ignition-delay:
      - 347 us
    pressure:
      - 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature:
      - 1389 kelvin
    ignition-delay:
      - 756 us
    pressure:
      - 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
    ...
\end{yamlbox}
%
As before, we have truncated the datapoints list for space.

@@

Then, we load the ChemKED file and generate a list of initial conditions that will be mapped onto
the \pybox|run_simulation()| function. We first define a convenience function to collect the input
from a single datapoint and return a Python tuple with the conditions (tuples are lightweight groups
of data in Python). Then, we use the built-in \pybox|map| function to apply the
\pybox|collect_input| function to each of the elements in the \pybox|ck.datapoints| list:
%
\begin{pythonbox}
ck = ChemKED('Stranic2012-tbuoh.yaml')

@@

initial_conditions = list(map(collect_input, ck.datapoints))
\end{pythonbox}

Finally, we create the processor \pybox|Pool| (with 4 processes) and send the jobs out to run:
%
\begin{pythonbox}
with Pool(processes=4) as pool:
    ignition_delays = pool.starmap(run_simulation, initial_conditions)
